---
title: "Machine Learning Foundations for Game Theory"
short-title: "ML Foundations"
author: "Raban Heller"
date-modified: today
abstract: "Supervised learning as a game against nature, regret minimization and regret matching, feature engineering for game-theoretic data, and cross-validation for strategic predictions."
keywords: ["machine learning", "regret matching", "supervised learning", "logistic regression", "cross-validation"]
---

# ML Foundations for Game Theory {#sec-ml-foundations}

```{r}
#| label: setup
#| include: false
source(here::here("R", "_common.R"))
```

## Learning objectives {.unnumbered}

- Frame supervised learning as a two-player game between a learner and nature.
- Implement regret matching from scratch and show convergence to Nash equilibrium in Rock-Paper-Scissors.
- Construct features from game-theoretic data suitable for prediction tasks.
- Apply cross-validation to evaluate predictive models of strategic behavior.

## Motivation

Machine learning and game theory share a deep intellectual kinship. At its core, supervised learning is a minimax problem: a learner selects a hypothesis to minimize worst-case loss, while "nature" (the data-generating process) selects inputs that expose the learner's weaknesses. This adversarial framing is not merely a metaphor — it is the foundation of online learning theory and connects directly to regret minimization, a concept central to modern algorithmic game theory.

This chapter bridges the two fields. We show how regret matching — a simple no-regret algorithm — converges to Nash equilibrium in two-player zero-sum games, then pivot to practical ML tools (logistic regression, cross-validation) applied to game-theoretic prediction tasks. The goal is to equip the reader with ML techniques that will appear throughout the remaining chapters on reinforcement learning, self-play, and counterfactual regret minimization.

## Theory

### Supervised learning as a game

In the **statistical learning** framework, a learner observes training data $\{(x_i, y_i)\}_{i=1}^n$ drawn from an unknown distribution $\mathcal{D}$ and selects a hypothesis $h$ from a class $\mathcal{H}$ to minimize expected loss:

$$
\min_{h \in \mathcal{H}} \; \mathbb{E}_{(x,y) \sim \mathcal{D}} \left[ \ell(h(x), y) \right]
$$ {#eq-erm}

This has a natural game-theoretic interpretation. The learner is Player 1, choosing $h$. Nature is Player 2, choosing $(x, y)$. The loss function $\ell$ is the payoff matrix — the learner minimizes it, nature maximizes it. The minimax theorem guarantees that in this zero-sum formulation, optimal learning strategies exist.

### Online learning and regret

In **online learning**, decisions are made sequentially. At each round $t$, the learner chooses an action $a_t$ from a set $\mathcal{A}$, then observes the loss vector $\ell_t(a)$ for all actions. The **cumulative regret** of not having played action $a$ is:

$$
R_T(a) = \sum_{t=1}^{T} \ell_t(a_t) - \sum_{t=1}^{T} \ell_t(a)
$$ {#eq-regret}

A **no-regret** algorithm ensures that $\max_a R_T(a) / T \to 0$ as $T \to \infty$. The celebrated connection to game theory is: if both players in a zero-sum game use no-regret algorithms, their time-averaged strategies converge to a Nash equilibrium [@cesa-bianchi2006].

### Regret matching

**Regret matching** [@hart2000] is one of the simplest no-regret algorithms. At each round, the player computes cumulative regret for each action and plays proportionally to positive regrets:

$$
\sigma_t(a) = \begin{cases}
R_t^+(a) \,/\, \sum_{a'} R_t^+(a') & \text{if } \sum_{a'} R_t^+(a') > 0 \\
1 / |\mathcal{A}| & \text{otherwise}
\end{cases}
$$ {#eq-regret-matching}

where $R_t^+(a) = \max(0, R_t(a))$. This algorithm is the building block of **counterfactual regret minimization** (CFR), covered in @sec-cfr-poker.

### Feature representation for game-theoretic data

When applying ML to strategic settings, feature engineering matters. Common features include:

- **History features**: opponent's action frequencies over a sliding window.
- **Payoff features**: expected payoff under each action given an opponent model.
- **State features**: round number, cumulative scores, cooperation rate.

## Implementation in R {#sec-ml-foundations-impl}

### Regret matching for Rock-Paper-Scissors

```{r}
#| label: regret-matching
regret_matching <- function(n_iterations = 5000) {
  actions <- c("Rock", "Paper", "Scissors")
  n_actions <- length(actions)
  # Payoff matrix for row player: rows = P1, cols = P2
  payoff <- matrix(c( 0, -1,  1,
                       1,  0, -1,
                      -1,  1,  0), nrow = 3, byrow = TRUE)

  cum_regret_1 <- rep(0, n_actions)
  cum_regret_2 <- rep(0, n_actions)
  cum_strategy_1 <- rep(0, n_actions)
  cum_strategy_2 <- rep(0, n_actions)

  strategy_history <- matrix(0, nrow = n_iterations, ncol = n_actions)

  get_strategy <- function(cum_regret) {
    pos_regret <- pmax(cum_regret, 0)
    total <- sum(pos_regret)
    if (total > 0) pos_regret / total else rep(1 / n_actions, n_actions)
  }

  for (t in seq_len(n_iterations)) {
    sigma_1 <- get_strategy(cum_regret_1)
    sigma_2 <- get_strategy(cum_regret_2)
    cum_strategy_1 <- cum_strategy_1 + sigma_1
    cum_strategy_2 <- cum_strategy_2 + sigma_2
    strategy_history[t, ] <- cum_strategy_1 / sum(cum_strategy_1)

    # Sample actions
    a1 <- sample(n_actions, 1, prob = sigma_1)
    a2 <- sample(n_actions, 1, prob = sigma_2)

    # Update regrets
    for (a in seq_len(n_actions)) {
      cum_regret_1[a] <- cum_regret_1[a] + payoff[a, a2] - payoff[a1, a2]
      cum_regret_2[a] <- cum_regret_2[a] + (-payoff[a1, a]) - (-payoff[a1, a2])
    }
  }

  avg_strategy_1 <- cum_strategy_1 / sum(cum_strategy_1)
  avg_strategy_2 <- cum_strategy_2 / sum(cum_strategy_2)

  list(
    avg_strategy_1 = setNames(avg_strategy_1, actions),
    avg_strategy_2 = setNames(avg_strategy_2, actions),
    history = strategy_history
  )
}
```

```{r}
#| label: run-regret-matching
set.seed(42)
rm_result <- regret_matching(n_iterations = 5000)
```

```{r}
#| label: fig-regret-convergence
#| fig-cap: "Regret matching convergence in Rock-Paper-Scissors. Player 1's time-averaged strategy converges toward the Nash equilibrium of (1/3, 1/3, 1/3) for each action."
#| fig-width: 7
#| fig-height: 5

rm_df <- tibble(
  iteration = rep(1:5000, 3),
  action = rep(c("Rock", "Paper", "Scissors"), each = 5000),
  probability = c(rm_result$history[, 1],
                  rm_result$history[, 2],
                  rm_result$history[, 3])
)

p_regret <- ggplot(rm_df, aes(x = iteration, y = probability, colour = action)) +
  geom_line(alpha = 0.8) +
  geom_hline(yintercept = 1/3, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("Rock" = okabe_ito[1],
                                  "Paper" = okabe_ito[2],
                                  "Scissors" = okabe_ito[3])) +
  theme_publication() +
  labs(title = "Regret Matching Convergence in RPS",
       x = "Iteration", y = "Average strategy probability",
       colour = "Action") +
  annotate("text", x = 4500, y = 0.36, label = "NE = 1/3",
           colour = "grey40", size = 3.5)

p_regret
save_pub_fig(p_regret, "regret-matching-convergence", width = 7, height = 5)
```

The average strategies converge to $(1/3, 1/3, 1/3)$ — the unique Nash equilibrium of Rock-Paper-Scissors. This confirms the theoretical guarantee: regret matching is a no-regret algorithm, and when both players use it, the time-averaged play converges to NE.

```{r}
#| label: rm-final-strategy
cat("Player 1 average strategy:\n")
print(round(rm_result$avg_strategy_1, 4))
cat("\nPlayer 2 average strategy:\n")
print(round(rm_result$avg_strategy_2, 4))
```

### Decision boundary visualization

To illustrate supervised learning in a game-theoretic context, consider a binary classification task: given two features derived from game play, predict an outcome. We fit a logistic regression model using `glm` and visualize the decision boundary.

```{r}
#| label: decision-boundary-data
set.seed(123)
n <- 200

# Simulate two player types: cooperators (y=1) and defectors (y=0)
# Feature 1: opponent's cooperation rate; Feature 2: round number (normalized)
x1_coop <- rnorm(n/2, mean = 0.7, sd = 0.15)
x2_coop <- rnorm(n/2, mean = 0.6, sd = 0.2)
x1_def  <- rnorm(n/2, mean = 0.3, sd = 0.15)
x2_def  <- rnorm(n/2, mean = 0.4, sd = 0.2)

game_data <- tibble(
  opp_coop_rate = c(x1_coop, x1_def),
  normalized_round = c(x2_coop, x2_def),
  cooperate = factor(c(rep(1, n/2), rep(0, n/2)))
)

model <- glm(cooperate ~ opp_coop_rate + normalized_round,
             data = game_data, family = binomial)
```

```{r}
#| label: fig-decision-boundary
#| fig-cap: "Decision boundary for predicting cooperation. The logistic regression separates cooperators (orange) from defectors (blue) based on the opponent's historical cooperation rate and normalized round number."
#| fig-width: 7
#| fig-height: 5

# Create prediction grid
grid <- expand.grid(
  opp_coop_rate = seq(-0.1, 1.1, length.out = 200),
  normalized_round = seq(-0.2, 1.2, length.out = 200)
)
grid$prob <- predict(model, newdata = grid, type = "response")

p_boundary <- ggplot() +
  geom_tile(data = grid, aes(x = opp_coop_rate, y = normalized_round,
                              fill = prob), alpha = 0.4) +
  scale_fill_gradient2(low = okabe_ito[2], mid = "white", high = okabe_ito[1],
                       midpoint = 0.5, name = "P(Cooperate)") +
  geom_point(data = game_data, aes(x = opp_coop_rate, y = normalized_round,
                                    colour = cooperate),
             size = 1.5, alpha = 0.7) +
  scale_colour_manual(values = c("0" = okabe_ito[2], "1" = okabe_ito[1]),
                      labels = c("Defect", "Cooperate"), name = "Outcome") +
  theme_publication() +
  labs(title = "Decision Boundary: Predicting Cooperation",
       x = "Opponent cooperation rate", y = "Normalized round")

p_boundary
save_pub_fig(p_boundary, "decision-boundary-cooperation", width = 7, height = 5)
```

## Worked example

### Predicting cooperation in the iterated Prisoner's Dilemma

We simulate an iterated Prisoner's Dilemma and use logistic regression to predict whether a player cooperates on a given round, based on observable features.

```{r}
#| label: ipd-simulation
set.seed(42)
n_rounds <- 200
n_pairs <- 50

# Simulate IPD data: each pair plays n_rounds
ipd_data <- map_dfr(seq_len(n_pairs), function(pair_id) {
  # Each player has a base cooperation probability that drifts with opponent behavior
  p1_coop <- 0.5
  p2_coop <- 0.5
  records <- vector("list", n_rounds)

  for (r in seq_len(n_rounds)) {
    a1 <- rbinom(1, 1, p1_coop)
    a2 <- rbinom(1, 1, p2_coop)
    # Tit-for-tat-ish adaptation
    p1_coop <- 0.3 * p1_coop + 0.7 * (0.2 + 0.6 * a2)
    p2_coop <- 0.3 * p2_coop + 0.7 * (0.2 + 0.6 * a1)

    records[[r]] <- tibble(pair = pair_id, round = r, p1_action = a1, p2_action = a2)
  }
  bind_rows(records)
})

# Rolling mean over previous 3 rounds (base R, no slider)
roll_mean3 <- function(x) {
  n <- length(x)
  out <- rep(NA_real_, n)
  for (i in 4:n) out[i] <- mean(x[(i - 3):(i - 1)])
  out
}

# Engineer features for predicting P1's cooperation
ipd_features <- ipd_data |>
  group_by(pair) |>
  mutate(
    opp_coop_last3 = roll_mean3(p2_action),
    own_coop_last3 = roll_mean3(p1_action),
    round_norm = round / n_rounds
  ) |>
  ungroup() |>
  filter(round > 4)
```

```{r}
#| label: ipd-model
# Split train/test by pair (avoid data leakage)
train_pairs <- sample(n_pairs, 35)
train <- ipd_features |> filter(pair %in% train_pairs)
test  <- ipd_features |> filter(!pair %in% train_pairs)

ipd_model <- glm(p1_action ~ opp_coop_last3 + own_coop_last3 + round_norm,
                 data = train, family = binomial)

cat("Logistic regression coefficients:\n")
print(round(coef(ipd_model), 3))
```

```{r}
#| label: ipd-cv
# 5-fold cross-validation by pair
folds <- split(train_pairs, cut(seq_along(train_pairs), 5, labels = FALSE))

cv_accuracy <- map_dbl(folds, function(held_out) {
  cv_train <- ipd_features |> filter(pair %in% setdiff(train_pairs, held_out))
  cv_test  <- ipd_features |> filter(pair %in% held_out)

  fit <- glm(p1_action ~ opp_coop_last3 + own_coop_last3 + round_norm,
             data = cv_train, family = binomial)

  preds <- ifelse(predict(fit, newdata = cv_test, type = "response") > 0.5, 1, 0)
  mean(preds == cv_test$p1_action)
})

cat(sprintf("5-fold CV accuracy: %.1f%% (SD: %.1f%%)\n",
            100 * mean(cv_accuracy), 100 * sd(cv_accuracy)))
```

```{r}
#| label: ipd-test-eval
test_preds <- predict(ipd_model, newdata = test, type = "response")
test_class <- ifelse(test_preds > 0.5, 1, 0)
test_accuracy <- mean(test_class == test$p1_action)

cat(sprintf("Test set accuracy: %.1f%%\n", 100 * test_accuracy))
cat(sprintf("Baseline (always predict majority class): %.1f%%\n",
            100 * max(mean(test$p1_action), 1 - mean(test$p1_action))))
```

The opponent's recent cooperation rate is the strongest predictor — consistent with tit-for-tat-like dynamics. Cross-validation by pair (rather than by round) prevents data leakage from correlated within-pair observations, a critical methodological point when applying ML to repeated games.

## Extensions

- **Multiplicative weights update** is an alternative no-regret algorithm with tighter theoretical bounds. See @arora2012 for a survey of the multiplicative weights method and its connections to game theory.
- **Counterfactual regret minimization** (@sec-cfr-poker) extends regret matching to extensive-form games, powering superhuman poker AI.
- **Reinforcement learning** (@sec-reinforcement-learning) replaces the fixed-loss setting of online learning with sequential decision-making under an MDP.
- **Feature engineering** for strategic data is an active research area: deep learning approaches learn features automatically, but interpretable features (cooperation rates, payoff summaries) remain valuable for analysis.

## Exercises {.unnumbered}

1. **Regret matching in Prisoner's Dilemma.** Implement regret matching for the one-shot Prisoner's Dilemma with payoffs $T = 5, R = 3, P = 1, S = 0$. To what strategy profile does the average play converge? Explain why this differs from RPS.

2. **Regularization and overfitting.** Add polynomial features (quadratic and interaction terms) to the IPD logistic regression model. Compare training and test accuracy. Does the richer model overfit? Use cross-validation to select the best model.

3. **Window size sensitivity.** In the worked example, features use a 3-round sliding window. Repeat the analysis with window sizes of 1, 5, 10, and 20. Plot cross-validation accuracy versus window size and explain the trade-off between responsiveness and stability.

Solutions appear in @sec-solutions.
