---
title: "LLM Agents and Strategic Reasoning"
short-title: "LLM Agents"
author: "Raban Heller"
date-modified: today
abstract: "Large language models as game-playing agents: level-k reasoning, cognitive hierarchy theory, bounded rationality, and simulated beauty contest experiments in R."
keywords: ["LLM", "language model", "agents", "strategic reasoning", "level-k", "cognitive hierarchy"]
---

# LLM Agents and Strategy {#sec-llm-agents}

```{r}
#| label: setup
#| include: false
source(here::here("R", "_common.R"))
```

## Learning objectives {.unnumbered}

- Describe how LLMs can serve as game-playing agents and the challenges of evaluating their strategic sophistication.
- Explain level-k reasoning and cognitive hierarchy theory as models of bounded rationality.
- Simulate a p-beauty contest with level-0 through level-3 players in R.
- Compare the strategic performance of different reasoning levels against Nash equilibrium predictions.

## Motivation

Large language models (LLMs) have shown surprising abilities in strategic settings. When prompted to play matrix games or auction settings, LLMs exhibit behavior that resembles --- but does not perfectly match --- game-theoretic rationality. Do they compute Nash equilibria? Do they reason about opponents? Or do they pattern-match from training data?

These questions connect directly to a long tradition in behavioral game theory: humans also deviate from Nash equilibrium in systematic ways. The **level-k** model [@stahl1995] and **cognitive hierarchy** theory [@camerer2004] were developed to explain human strategic behavior. These same frameworks provide a lens for understanding LLM agents.

This chapter is more conceptual than implementation-heavy, since we cannot call LLM APIs from our R setup. Instead, we build the theoretical framework and simulate the models that researchers use to evaluate strategic sophistication in both humans and AI agents.

## Theory

### LLMs as game-playing agents

An LLM can play a game when given a natural-language description of the rules, payoffs, and its role. The "strategy" emerges from the model's text generation process rather than from explicit optimization. Several empirical findings have emerged:

- **Matrix games**: LLMs often cooperate in Prisoner's Dilemma variants at rates similar to human subjects, but their behavior is sensitive to framing effects (how the game is described).
- **Coordination games**: LLMs can coordinate on focal points, suggesting they leverage the same cultural knowledge that helps humans coordinate.
- **Auctions and bargaining**: LLMs show varying degrees of strategic sophistication, sometimes overbidding or underbidding relative to equilibrium predictions.

### Level-k reasoning

The **level-k** model posits that players differ in their depth of strategic reasoning:

- **Level-0** ($L_0$): Plays uniformly at random (or according to some non-strategic rule).
- **Level-1** ($L_1$): Best-responds to a level-0 opponent.
- **Level-2** ($L_2$): Best-responds to a level-1 opponent.
- **Level-k** ($L_k$): Best-responds to a level-$(k-1)$ opponent.

Each level assumes the opponent is exactly one level below. This anchored reasoning chain was proposed by @stahl1995 and contrasts with Nash equilibrium, which requires players to reason about opponents of the *same* sophistication level.

### Cognitive hierarchy

The **cognitive hierarchy** model [@camerer2004] generalizes level-k reasoning. Instead of assuming the opponent is exactly one level below, a level-$k$ player believes opponents are distributed across levels $0, 1, \ldots, k-1$ according to a Poisson distribution with parameter $\tau$:

$$
f(k) = \frac{e^{-\tau} \tau^k}{k!}
$$ {#eq-cognitive-hierarchy}

A level-$k$ player then best-responds to the *mixture* of lower-level strategies, weighted by their perceived frequencies. The parameter $\tau$ captures the average depth of reasoning in the population.

### The p-beauty contest

The **p-beauty contest** (or "guessing game") introduced by @nagel1995 is the canonical testbed for level-k reasoning. The rules:

1. $N$ players simultaneously choose a number from 0 to 100.
2. The winner is the player whose number is closest to $p$ times the average of all chosen numbers (typically $p = 2/3$).

The Nash equilibrium is for all players to choose 0 (by iterated elimination of dominated strategies). But in experiments, humans cluster around level-1 ($\approx 33$) and level-2 ($\approx 22$) responses.

### Bounded rationality and LLMs

LLMs exhibit a form of **bounded rationality**: they reason strategically but imperfectly, much like human subjects. Key parallels include:

- **Finite depth of reasoning**: LLMs rarely compute Nash equilibria from first principles; they appear to perform a few steps of iterated reasoning.
- **Framing effects**: The same game described differently elicits different strategies --- a hallmark of bounded rationality.
- **Anchoring**: LLMs, like humans, can anchor on salient numbers or focal points.

The level-k framework provides a natural benchmark: we can classify an LLM's strategic sophistication by determining which level-k model best fits its behavior across a battery of games.

## Implementation in R {#sec-llm-impl}

### Level-k beauty contest simulation

```{r}
#| label: beauty-contest
beauty_contest <- function(n_players = 20, p = 2/3, tau = 1.5,
                            level_distribution = NULL) {
  if (is.null(level_distribution)) {
    # Poisson cognitive hierarchy
    max_level <- 5
    levels <- 0:max_level
    probs <- dpois(levels, lambda = tau)
    probs <- probs / sum(probs)
    player_levels <- sample(levels, n_players, replace = TRUE, prob = probs)
  } else {
    player_levels <- sample(names(level_distribution), n_players,
                            replace = TRUE, prob = level_distribution)
    player_levels <- as.integer(player_levels)
  }

  # Compute level-k choices
  level_choice <- function(k, p) {
    # Level-0: uniform random -> expected value = 50
    # Level-k: best respond to level-(k-1)
    if (k == 0) return(50)
    50 * p^k
  }

  choices <- sapply(player_levels, level_choice, p = p)

  # Add small noise for realism
  choices <- pmax(0, pmin(100, choices + rnorm(n_players, 0, 2)))

  tibble(
    player = seq_len(n_players),
    level = player_levels,
    choice = choices
  )
}
```

### Run the simulation

```{r}
#| label: run-beauty-contest
set.seed(42)
n_rounds <- 500
p <- 2/3

results <- map_dfr(seq_len(n_rounds), function(round) {
  game <- beauty_contest(n_players = 20, p = p, tau = 1.5)
  target <- p * mean(game$choice)
  game |>
    mutate(
      round = round,
      target = target,
      distance = abs(choice - target),
      won = distance == min(distance)
    )
})
```

### Figure 1: Level-k action distributions

```{r}
#| label: fig-levelk-distributions
#| fig-cap: "Predicted choice distributions in a 2/3-beauty contest for different levels of reasoning. Level-0 players choose near 50 (uniform random anchor). Each subsequent level best-responds to the level below, producing geometrically decreasing choices that converge toward the Nash equilibrium of 0."
#| fig-width: 7
#| fig-height: 5

# Theoretical level-k distributions
level_data <- tibble(
  level = 0:4,
  label = paste0("Level-", 0:4),
  theoretical_choice = 50 * p^(0:4)
)

# Simulated distributions from our runs
sim_dist <- results |>
  filter(level <= 4) |>
  mutate(label = paste0("Level-", level))

p_levels <- ggplot(sim_dist, aes(x = choice, fill = label)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  geom_vline(data = level_data, aes(xintercept = theoretical_choice,
                                      colour = label),
             linetype = "dashed", linewidth = 0.8) +
  scale_fill_manual(values = setNames(okabe_ito[1:5],
                                       paste0("Level-", 0:4))) +
  scale_colour_manual(values = setNames(okabe_ito[1:5],
                                         paste0("Level-", 0:4))) +
  facet_wrap(~label, ncol = 1, scales = "free_y") +
  theme_publication() +
  theme(legend.position = "none") +
  labs(title = "Choice Distributions by Reasoning Level",
       subtitle = "p-Beauty Contest with p = 2/3",
       x = "Choice (0-100)", y = "Count") +
  xlim(0, 70)

p_levels
save_pub_fig(p_levels, "levelk-distributions", width = 7, height = 5)
```

### Figure 2: Strategy performance comparison

```{r}
#| label: fig-strategy-performance
#| fig-cap: "Win rates by strategy type in simulated beauty contest games. Level-1 and level-2 players outperform level-0 (random) players, but higher levels show diminishing returns because the population contains mostly low-level reasoners. The Nash equilibrium strategy (choose 0) rarely wins against bounded rational opponents."
#| fig-width: 7
#| fig-height: 4.5

# Compute win rates by level
win_rates <- results |>
  filter(level <= 4) |>
  group_by(level) |>
  summarise(
    n_games = n(),
    n_wins = sum(won),
    win_rate = n_wins / n_games,
    .groups = "drop"
  ) |>
  mutate(label = paste0("Level-", level))

# Add Nash equilibrium player for comparison
# Simulate NE player (choice = 0) against the population
set.seed(42)
ne_wins <- 0
ne_games <- n_rounds
for (r in seq_len(ne_games)) {
  game <- beauty_contest(n_players = 19, p = p, tau = 1.5)
  all_choices <- c(0, game$choice)  # NE player chooses 0

  target <- p * mean(all_choices)
  distances <- abs(all_choices - target)
  if (which.min(distances) == 1) ne_wins <- ne_wins + 1
}

# Add random player for comparison
set.seed(42)
rand_wins <- 0
for (r in seq_len(ne_games)) {
  game <- beauty_contest(n_players = 19, p = p, tau = 1.5)
  rand_choice <- runif(1, 0, 100)
  all_choices <- c(rand_choice, game$choice)
  target <- p * mean(all_choices)
  distances <- abs(all_choices - target)
  if (which.min(distances) == 1) rand_wins <- rand_wins + 1
}

comparison <- bind_rows(
  win_rates |> select(label, win_rate),
  tibble(label = "Nash (0)", win_rate = ne_wins / ne_games),
  tibble(label = "Random", win_rate = rand_wins / ne_games)
) |>
  mutate(label = factor(label,
    levels = c("Random", "Level-0", "Level-1", "Level-2",
               "Level-3", "Level-4", "Nash (0)")))

p_perf <- ggplot(comparison, aes(x = label, y = win_rate, fill = label)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 1/20, linetype = "dashed", colour = "grey50") +
  scale_fill_manual(values = c("Random" = "grey60", okabe_ito[1:5],
                                "Nash (0)" = okabe_ito[8])) +
  theme_publication() +
  theme(legend.position = "none") +
  labs(title = "Win Rates by Strategy Type",
       subtitle = "p-Beauty Contest (p = 2/3, 20 players, Poisson CH population)",
       x = "Strategy", y = "Win rate") +
  annotate("text", x = 6.5, y = 1/20 + 0.005, label = "Chance = 1/20",
           colour = "grey50", size = 3.5)

p_perf
save_pub_fig(p_perf, "strategy-performance-comparison", width = 7, height = 4.5)
```

## Worked example

We walk through a single round of the p-beauty contest with explicit level-k reasoning.

```{r}
#| label: worked-example
cat("=== p-Beauty Contest Walkthrough (p = 2/3) ===\n\n")

cat("Step 1: Level-0 reasoning\n")
cat("  A level-0 player chooses randomly; expected value = 50.\n\n")

cat("Step 2: Level-1 reasoning\n")
cat("  'If everyone else is level-0, the average will be ~50.'\n")
cat(sprintf("  Best response: p * 50 = 2/3 * 50 = %.1f\n\n", p * 50))

cat("Step 3: Level-2 reasoning\n")
cat("  'If everyone else is level-1, the average will be ~33.3.'\n")
cat(sprintf("  Best response: p * 33.3 = 2/3 * 33.3 = %.1f\n\n", p^2 * 50))

cat("Step 4: Level-3 reasoning\n")
cat("  'If everyone else is level-2, the average will be ~22.2.'\n")
cat(sprintf("  Best response: p * 22.2 = 2/3 * 22.2 = %.1f\n\n", p^3 * 50))

cat("Step 5: Nash equilibrium (infinite levels)\n")
cat("  As k -> infinity, the choice converges to 0.\n")
cat("  But choosing 0 only wins if ALL other players also choose 0.\n\n")

# Show a concrete simulated round
set.seed(123)
example_game <- beauty_contest(n_players = 10, p = p, tau = 1.5)
target <- p * mean(example_game$choice)

cat("=== Example Round with 10 Players ===\n\n")
example_game |>
  mutate(
    label = paste0("Level-", level),
    distance = abs(choice - target)
  ) |>
  arrange(distance) |>
  select(player, label, choice, distance) |>
  rename(Player = player, Level = label,
         Choice = choice, `Dist to target` = distance) |>
  print(n = 10)

cat(sprintf("\nAverage choice: %.1f\n", mean(example_game$choice)))
cat(sprintf("Target (2/3 * average): %.1f\n", target))
cat(sprintf("Winner: Player %d (Level-%d, chose %.1f)\n",
            example_game$player[which.min(abs(example_game$choice - target))],
            example_game$level[which.min(abs(example_game$choice - target))],
            example_game$choice[which.min(abs(example_game$choice - target))]))
```

**Key takeaway.** In a population of bounded rational players, the Nash equilibrium (choose 0) performs poorly. The best strategy depends on the sophistication distribution of the population --- exactly the insight from cognitive hierarchy theory. An LLM that reasons at approximately level-2 would outperform both a naive randomizer and a fully rational Nash player in this setting.

## Extensions

- **Experimental design for LLM evaluation**: To test an LLM's strategic level, present it with multiple beauty contest games using different $p$ values and population descriptions. A consistent level-k response pattern reveals the model's reasoning depth.
- **Prompt sensitivity**: Rephrasing the same game (e.g., "competitive" vs. "cooperative" framing) can shift LLM behavior by 1--2 levels, paralleling human framing effects.
- **Multi-game batteries**: Testing across Prisoner's Dilemma, public goods games, and ultimatum games builds a richer profile of strategic sophistication.
- **Q-learning** (@sec-q-learning) and **multi-agent RL** (@sec-multi-agent-rl) offer alternative approaches to AI game-playing that learn through experience rather than language understanding.

For the foundational level-k theory, see @stahl1995; for cognitive hierarchy theory, see @camerer2004; and for the original beauty contest experiment, see @nagel1995.

## Exercises {.unnumbered}

1. **Varying $\tau$.** Simulate the beauty contest with cognitive hierarchy parameter $\tau \in \{0.5, 1.0, 1.5, 2.0, 3.0\}$. How does the average depth of reasoning affect the optimal level-k strategy? Plot win rates for each level as a function of $\tau$.

2. **Different $p$ values.** Run the beauty contest with $p \in \{1/3, 1/2, 2/3, 3/4\}$. How does the target multiplier affect the separation between level-k predictions? At what value of $p$ is level-1 vs. level-2 distinction largest?

3. **Iterated beauty contest.** Simulate 20 rounds of the beauty contest where players update their level each round: after each round, a player whose choice was far from the target increases their level by 1 (up to level 5). Does the population converge to Nash equilibrium over time? Plot the average choice per round.

Solutions appear in @sec-solutions.
