# (APPENDIX) Appendices {-}

# R Refresher {#sec-r-refresher}

> A concise refresher on R fundamentals for readers who need a quick review.
```{r, setup, include=FALSE}
source(here::here("R", "_common.R"))
```

This appendix provides a quick reference for the R features used most often in this book. It is not a substitute for a full R tutorial but should be enough to remind you of the essentials. For a deeper treatment, see @wickham2023r.

## Data structures

### Vectors

Vectors are the fundamental data type in R. Every element must share the same type (logical, integer, double, character).

```{r, vectors}
# Numeric vector
payoffs <- c(3, 0, 5, 1)

# Character vector
strategies <- c("Cooperate", "Defect")

# Logical vector
is_dominated <- c(FALSE, TRUE, FALSE)

# Sequence shortcuts
indices <- 1:10
grid <- seq(0, 1, by = 0.1)

# Vectorised arithmetic
payoffs * 2
payoffs > 2
```

### Matrices

Matrices are two-dimensional vectors. They are central to payoff representations throughout this book.

```{r, matrices}
# Create a 2x2 payoff matrix (filled by column by default)
A <- matrix(c(3, 5, 0, 1), nrow = 2, byrow = TRUE)
rownames(A) <- c("Cooperate", "Defect")
colnames(A) <- c("Cooperate", "Defect")
A

# Matrix indexing
A[1, 2]       # row 1, column 2
A["Defect", ] # entire row by name

# Dimensions
dim(A)
nrow(A)
ncol(A)
```

### Lists

Lists can hold elements of different types and different lengths. They are used throughout the book to bundle game components together.

```{r, lists}
game <- list(
  players  = c("Player 1", "Player 2"),
  payoff_1 = matrix(c(3, 5, 0, 1), nrow = 2, byrow = TRUE),
  payoff_2 = matrix(c(3, 0, 5, 1), nrow = 2, byrow = TRUE)
)

# Access by name
game$players
game[["payoff_1"]]

# Access by position
game[[1]]
```

### Data frames and tibbles

Data frames (and their tidyverse counterpart, tibbles) are the primary structure for tabular data.

```{r, tibbles}
# Base R data frame
df <- data.frame(
  strategy = c("Cooperate", "Defect"),
  payoff   = c(3, 1)
)

# Tidyverse tibble
tbl <- tibble(
  strategy = c("Cooperate", "Defect"),
  payoff   = c(3, 1)
)

# Tibbles print more informatively
tbl
```

## Control flow

### Conditionals

```{r, conditionals}
x <- 5

if (x > 3) {
  result <- "high"
} else if (x > 1) {
  result <- "medium"
} else {
  result <- "low"
}
result

# Vectorised conditional
ifelse(c(3, 0, 5, 1) > 2, "above", "below")
```

### Loops

```{r, loops}
# for loop
total <- 0
for (i in 1:5) {
  total <- total + i
}
total

# while loop
count <- 0
value <- 1
while (value < 100) {
  value <- value * 2
  count <- count + 1
}
count
```

### The apply family

The `apply` functions replace explicit loops with concise functional calls.

```{r, apply}
A <- matrix(1:12, nrow = 3)

# Apply across rows (margin = 1)
apply(A, 1, sum)

# Apply across columns (margin = 2)
apply(A, 2, mean)

# sapply returns a vector; lapply returns a list
sapply(1:5, function(x) x^2)
lapply(1:3, function(x) rep(x, x))

# vapply is like sapply but with a specified return type
vapply(1:5, function(x) x^2, numeric(1))
```

## Functions

### Defining functions

```{r, functions}
# A function that computes expected payoff
expected_payoff <- function(payoff_vec, prob_vec) {
  sum(payoff_vec * prob_vec)
}

expected_payoff(c(3, 0), c(0.6, 0.4))

# Default arguments
greet <- function(name, greeting = "Hello") {
  paste(greeting, name)
}
greet("Alice")
greet("Bob", greeting = "Hi")
```

### Closures and environments

A closure is a function that captures variables from its enclosing environment. This pattern is useful for creating parameterised game constructors.

```{r, closures}
make_discounter <- function(delta) {
  # delta is captured in the closure
  function(payoffs) {
    n <- length(payoffs)
    weights <- delta^(seq_len(n) - 1)
    sum(payoffs * weights)
  }
}

discount_95 <- make_discounter(0.95)
discount_95(c(3, 3, 3, 3, 3))

# The enclosed value of delta persists
discount_50 <- make_discounter(0.50)
discount_50(c(3, 3, 3, 3, 3))
```

### Anonymous functions

R 4.1+ supports the shorthand `\(x)` syntax for anonymous (lambda) functions.

```{r, anonymous}
sapply(1:5, \(x) x^2)

# Equivalent to the older syntax:
sapply(1:5, function(x) x^2)
```

## Tidyverse basics

The tidyverse is a collection of packages for data wrangling and visualisation. This book uses it extensively.

### The pipe operator

The pipe `|>` (base R 4.1+) or `%>%` (magrittr) passes the left-hand side as the first argument to the right-hand side.

```{r, pipe}
c(4, 1, 7, 2, 9) |>
  sort() |>
  rev() |>
  head(3)
```

### dplyr verbs

The five core verbs handle the vast majority of data manipulation tasks.

```{r, dplyr}
game_results <- tibble(
  round    = 1:6,
  player_1 = c("C", "D", "C", "C", "D", "D"),
  player_2 = c("C", "C", "D", "C", "D", "C"),
  payoff_1 = c(3, 5, 0, 3, 1, 5),
  payoff_2 = c(3, 0, 5, 3, 1, 0)
)

# filter: keep rows matching a condition
game_results |> filter(player_1 == "C")

# mutate: create or modify columns
game_results |> mutate(total = payoff_1 + payoff_2)

# select: choose columns
game_results |> select(round, payoff_1, payoff_2)

# arrange: reorder rows
game_results |> arrange(desc(payoff_1))

# summarise (often with group_by)
game_results |>
  group_by(player_1) |>
  summarise(
    mean_payoff = mean(payoff_1),
    n_rounds    = n()
  )
```

### Pivoting

Converting between wide and long formats is essential for plotting.

```{r, pivot}
# Wide to long
game_long <- game_results |>
  pivot_longer(
    cols      = starts_with("payoff"),
    names_to  = "player",
    values_to = "payoff"
  )
head(game_long)

# Long to wide
game_long |>
  pivot_wider(names_from = player, values_from = payoff)
```

### ggplot2 layers

The grammar of graphics builds plots layer by layer: data, aesthetics, geometries, and scales.
```{r, ggplot2-example, fig.cap="Payoffs across six rounds of a repeated game."}
game_results |>
  pivot_longer(
    cols = starts_with("payoff"),
    names_to = "player",
    values_to = "payoff"
  ) |>
  ggplot(aes(x = round, y = payoff, colour = player)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_colour_manual(
    values = c(payoff_1 = "#1b9e77", payoff_2 = "#d95f02"),
    labels = c("Player 1", "Player 2")
  ) +
  labs(x = "Round", y = "Payoff", colour = NULL) +
  theme_minimal()
```

## Useful idioms for this book

A handful of patterns recur throughout the chapters.

```{r, idioms}
# Replicate a simulation many times
results <- replicate(1000, {
  sample(c("Heads", "Tails"), 1)
})
table(results)

# Outer product for payoff grids
p <- seq(0, 1, by = 0.25)
q <- seq(0, 1, by = 0.25)
payoff_grid <- outer(p, q, function(pi, qi) 3 * pi * qi + 1 * (1 - pi))
payoff_grid

# Named vectors for readable code
strategy_names <- c(C = "Cooperate", D = "Defect")
strategy_names["C"]

# Setting seeds for reproducibility
set.seed(42)
sample(1:100, 5)
```
