# Probability Essentials {#sec-probability}

> Probability distributions, expectations, Bayes' rule, and stochastic processes.
```{r, setup, include=FALSE}
source(here::here("R", "_common.R"))
```

This appendix reviews the probability concepts that underpin mixed strategies, Bayesian games, and the simulation chapters. The emphasis is on intuition and R implementation rather than measure-theoretic rigour.

## Sample spaces and events

A **sample space** $\Omega$ is the set of all possible outcomes of a random experiment. An **event** $E \subseteq \Omega$ is a subset of outcomes. A **probability measure** $P$ assigns a number $P(E) \in [0, 1]$ to every event, with $P(\Omega) = 1$.

In game theory, the sample space often corresponds to the set of action profiles or the set of player types.

## Discrete distributions

A **discrete random variable** $X$ takes values in a countable set $\{x_1, x_2, \ldots\}$ with probability mass function $p(x_i) = P(X = x_i)$.

### Common discrete distributions

```{r, discrete-dists}
# Bernoulli: coin flip (used in mixed strategies)
rbinom(10, size = 1, prob = 0.5)

# Binomial: number of cooperators in n rounds
dbinom(3, size = 10, prob = 0.6)  # P(X = 3)

# Uniform: equally likely strategies
sample(c("Rock", "Paper", "Scissors"), size = 1)

# Poisson: arrival counts (used in some mechanism design models)
rpois(10, lambda = 3)
```

### Probability mass function and CDF

```{r, pmf-cdf}
# PMF and CDF of Binomial(10, 0.4)
x <- 0:10
pmf <- dbinom(x, size = 10, prob = 0.4)
cdf <- pbinom(x, size = 10, prob = 0.4)

tibble(x = x, pmf = pmf, cdf = cdf) |>
  head(6)
```

## Continuous distributions

A **continuous random variable** $X$ has a probability density function $f(x)$ such that $P(a \leq X \leq b) = \int_a^b f(x)\, dx$.

### Common continuous distributions

```{r, continuous-dists}
# Uniform on [0, 1]: used for types in Bayesian games
runif(5, min = 0, max = 1)

# Normal: payoff noise, error terms
rnorm(5, mean = 0, sd = 1)

# Exponential: waiting times, discount factors
rexp(5, rate = 1)
```

### Density, CDF, and quantiles

R uses a consistent naming convention: `d` for density, `p` for CDF, `q` for quantile, and `r` for random draws.

```{r, dpqr}
# Normal distribution
dnorm(0, mean = 0, sd = 1)   # density at 0
pnorm(1.96, mean = 0, sd = 1) # P(X <= 1.96)
qnorm(0.975, mean = 0, sd = 1) # 97.5th percentile
rnorm(3, mean = 0, sd = 1)    # three random draws
```

## Expected value and variance

The **expected value** of a discrete random variable is $E[X] = \sum_i x_i \, p(x_i)$. For a continuous variable, $E[X] = \int x \, f(x) \, dx$. In game theory, the expected payoff under a mixed strategy $\mathbf{p}$ is:

$$
E[u_i] = \sum_{a \in A} p(a) \, u_i(a)
$$

The **variance** $\text{Var}(X) = E[(X - E[X])^2] = E[X^2] - (E[X])^2$ measures the spread of outcomes.

```{r, expected-value}
# Expected payoff under a mixed strategy
payoffs <- c(3, 0, 5, 1)
probs   <- c(0.3, 0.2, 0.1, 0.4)

# Expected value
ev <- sum(payoffs * probs)
ev

# Variance
variance <- sum(probs * (payoffs - ev)^2)
variance

# Using simulation to verify
simulated <- sample(payoffs, size = 100000, replace = TRUE, prob = probs)
mean(simulated)
var(simulated)
```

### Linearity of expectation

A key property: $E[aX + bY] = aE[X] + bE[Y]$, regardless of whether $X$ and $Y$ are independent. This is used constantly when computing expected payoffs in multi-player games.

## Conditional probability

The **conditional probability** of $A$ given $B$ is:

$$
P(A \mid B) = \frac{P(A \cap B)}{P(B)}, \quad P(B) > 0
$$

Two events are **independent** if $P(A \cap B) = P(A) P(B)$, equivalently $P(A \mid B) = P(A)$.

```{r, conditional}
# Simulation: P(both cooperate | Player 1 cooperates)
n <- 100000
p1 <- sample(c("C", "D"), n, replace = TRUE, prob = c(0.6, 0.4))
p2 <- sample(c("C", "D"), n, replace = TRUE, prob = c(0.5, 0.5))

# Joint and conditional (assuming independence here)
p_both_C <- mean(p1 == "C" & p2 == "C")
p_p1_C   <- mean(p1 == "C")
p_both_C / p_p1_C  # should be approx 0.5
```

## Bayes' theorem

**Bayes' theorem** is the foundation of Bayesian games (\@ref(sec-bayesian-games)). It tells us how to update beliefs about a player's type after observing their action:

$$
P(\theta \mid a) = \frac{P(a \mid \theta) \, P(\theta)}{P(a)}
$$

where $\theta$ is the player's type, $a$ is the observed action, $P(\theta)$ is the prior, $P(a \mid \theta)$ is the likelihood, and $P(a) = \sum_{\theta'} P(a \mid \theta') P(\theta')$ is the marginal likelihood.

```{r, bayes}
# Signaling game: worker chooses Education or No Education
# Types: High (prob 0.6), Low (prob 0.4)
# P(Education | High) = 0.9, P(Education | Low) = 0.2

prior_high <- 0.6
prior_low  <- 0.4

p_edu_given_high <- 0.9
p_edu_given_low  <- 0.2

# P(Education)
p_edu <- p_edu_given_high * prior_high + p_edu_given_low * prior_low
p_edu

# P(High | Education) by Bayes' theorem
p_high_given_edu <- (p_edu_given_high * prior_high) / p_edu
p_high_given_edu

# P(High | No Education)
p_no_edu <- 1 - p_edu
p_high_given_no_edu <- ((1 - p_edu_given_high) * prior_high) / p_no_edu
p_high_given_no_edu
```

### Sequential updating

When multiple signals are observed, Bayes' theorem can be applied sequentially: the posterior from one update becomes the prior for the next. This is used in repeated Bayesian games where players learn about opponents over time.

```{r, sequential-bayes}
# Sequential Bayesian updating: observe actions round by round
# Prior: P(Cooperator type) = 0.5
# P(C action | Cooperator type) = 0.9
# P(C action | Defector type) = 0.1

update_belief <- function(prior, likelihood_coop, likelihood_defect, action) {
  if (action == "C") {
    l_coop <- likelihood_coop
    l_defect <- likelihood_defect
  } else {
    l_coop <- 1 - likelihood_coop
    l_defect <- 1 - likelihood_defect
  }
  numerator   <- l_coop * prior
  denominator <- l_coop * prior + l_defect * (1 - prior)
  numerator / denominator
}

prior <- 0.5
actions <- c("C", "C", "D", "C", "C")

beliefs <- numeric(length(actions))
for (i in seq_along(actions)) {
  prior <- update_belief(prior, 0.9, 0.1, actions[i])
  beliefs[i] <- prior
}

tibble(round = seq_along(actions), action = actions, belief = beliefs)
```

## Law of large numbers and Monte Carlo

The **law of large numbers** guarantees that the sample mean converges to the true mean as the sample size grows. This justifies the Monte Carlo methods used extensively in \@ref(sec-monte-carlo).

```{r, lln}
# Demonstrate convergence of sample mean
n_values <- c(10, 100, 1000, 10000, 100000)
true_mean <- 3.5  # E[Uniform die roll]

estimates <- vapply(n_values, function(n) {
  mean(sample(1:6, n, replace = TRUE))
}, numeric(1))

tibble(n = n_values, estimate = estimates, true_mean = true_mean)
```

### Monte Carlo estimation

To estimate a quantity $\theta = E[g(X)]$, draw $N$ independent samples $X_1, \ldots, X_N$ and compute:

$$
\hat{\theta}_N = \frac{1}{N} \sum_{i=1}^{N} g(X_i)
$$

The standard error of this estimate is $\text{SE} = \sigma / \sqrt{N}$, where $\sigma$ is the standard deviation of $g(X)$.

```{r, monte-carlo}
# Estimate P(at least one cooperator in 5 players) via Monte Carlo
# Each player cooperates independently with probability 0.3

estimate_prob <- function(n_sims, n_players = 5, p_coop = 0.3) {
  counts <- replicate(n_sims, {
    actions <- rbinom(n_players, size = 1, prob = p_coop)
    as.integer(sum(actions) >= 1)
  })
  c(estimate = mean(counts), se = sd(counts) / sqrt(n_sims))
}

# Exact answer: 1 - (1 - 0.3)^5
exact <- 1 - 0.7^5

set.seed(42)
mc <- estimate_prob(100000)
tibble(
  method   = c("Monte Carlo", "Exact"),
  estimate = c(mc["estimate"], exact)
)
```

## Random sampling for games

Several chapters use random sampling to simulate strategic interactions. The key functions are:

```{r, sampling}
# sample(): draw from a discrete set
sample(c("Hawk", "Dove"), size = 10, replace = TRUE, prob = c(0.4, 0.6))

# replicate(): repeat a simulation
replicate(5, sum(sample(1:6, 2, replace = TRUE)))

# set.seed(): ensure reproducibility
set.seed(123)
runif(3)
set.seed(123)
runif(3)  # identical output
```

## Summary of R probability functions

| Distribution | PMF/PDF | CDF | Quantile | Random |
|:-------------|:--------|:----|:---------|:-------|
| Binomial | `dbinom()` | `pbinom()` | `qbinom()` | `rbinom()` |
| Poisson | `dpois()` | `ppois()` | `qpois()` | `rpois()` |
| Uniform | `dunif()` | `punif()` | `qunif()` | `runif()` |
| Normal | `dnorm()` | `pnorm()` | `qnorm()` | `rnorm()` |
| Exponential | `dexp()` | `pexp()` | `qexp()` | `rexp()` |

: R functions for common probability distributions {#tbl-prob-functions}
