---
title: "Chapter Title Here"
short-title: "Short Title"
author: "Raban Heller"
date-modified: today
abstract: "One-sentence summary of the chapter's contribution."
keywords: ["keyword1", "keyword2", "keyword3"]
---

# Chapter Title {#sec-chapter-slug}

```{r}
#| label: setup
#| include: false
source(here::here("R", "_common.R"))
```

## Learning objectives {.unnumbered}

- Objective 1
- Objective 2
- Objective 3

## Motivation

Concrete problem, why it matters, real-world example that hooks the reader.

## Theory

Definitions, theorems with citations [@citation-key], and LaTeX equations.

$$
\sigma_i^* \in \arg\max_{\sigma_i \in \Delta(A_i)} u_i(\sigma_i, \sigma_{-i}^*)
$$ {#eq-example}

As shown in @eq-example, the equilibrium condition requires...

::: {.callout-note}
## Definition
A formal definition block for key concepts.
:::

## Implementation in R

```{r}
#| label: fig-example-plot
#| fig-cap: "Description of what the figure shows."
# 30–60 lines of runnable R code
p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(colour = okabe_ito[1]) +
  theme_publication() +
  labs(title = "Example Figure", x = "Weight", y = "MPG")
p
save_pub_fig(p, "example-plot")
```

## Worked example

Step-by-step walkthrough with output shown, connecting theory to code.

## Extensions

Pointers to further reading and related chapters via cross-references
(e.g., @sec-chapter-slug).

## Exercises {.unnumbered}

1. Exercise 1 (conceptual).
2. Exercise 2 (computational — requires R).
3. Exercise 3 (extension — open-ended).

Solutions appear in @sec-solutions.
