---
title: "Linear Algebra Essentials"
short-title: "Linear Algebra"
author: "Raban Heller"
date-modified: today
abstract: "Vectors, matrices, eigenvalues, and the linear algebra used throughout this book."
keywords: ["linear algebra", "matrices", "eigenvalues", "vectors"]
other-links:
  - text: Download Chapter PDF
    href: ../downloads/linear-algebra-essentials.pdf
    icon: file-pdf
---
# Linear Algebra Essentials {#sec-linear-algebra}

```{r}
#| label: setup
#| include: false
source(here::here("R", "_common.R"))
```

This appendix reviews the linear algebra that appears most often in game theory. The focus is on concepts rather than proofs, with R implementations alongside each topic. For a thorough mathematical treatment, see @strang2022.

## Vectors and vector operations

A vector in $\mathbb{R}^n$ is an ordered list of $n$ real numbers. In game theory, vectors represent strategy profiles, probability distributions over actions, and payoff allocations.

```{r}
#| label: vector-ops
# Define vectors
x <- c(3, 1, 4)
y <- c(2, 7, 1)

# Scalar multiplication and addition
2 * x
x + y

# Dot (inner) product
sum(x * y)
# Or equivalently:
x %*% y

# Euclidean norm (length)
sqrt(sum(x^2))
```

The **dot product** $\mathbf{x} \cdot \mathbf{y} = \sum_{i} x_i y_i$ appears whenever we compute expected payoffs: if $\mathbf{p}$ is a probability vector and $\mathbf{u}$ is a payoff vector, then $\mathbf{p} \cdot \mathbf{u}$ is the expected payoff.

## Matrices

A matrix $A \in \mathbb{R}^{m \times n}$ is a rectangular array of numbers. In game theory, the primary use is the payoff matrix of a normal-form game.

### Matrix arithmetic in R

```{r}
#| label: matrix-arith
A <- matrix(c(3, 0, 5, 1), nrow = 2, byrow = TRUE)
B <- matrix(c(3, 5, 0, 1), nrow = 2, byrow = TRUE)

# Element-wise operations
A + B
A * B  # Hadamard (element-wise) product

# Matrix multiplication
A %*% B

# Transpose
t(A)
```

### Key matrix operations

```{r}
#| label: matrix-ops
M <- matrix(c(4, 2, 1, 3), nrow = 2, byrow = TRUE)

# Determinant
det(M)

# Inverse (only if det != 0)
solve(M)

# Verify: M %*% M^{-1} = I
M %*% solve(M)
```

## Systems of linear equations

Many game-theoretic computations reduce to solving a system $A\mathbf{x} = \mathbf{b}$. For example, finding a mixed Nash equilibrium in a 2x2 game requires solving the indifference equations (see @sec-mixed-strategies).

### Solving $A\mathbf{x} = \mathbf{b}$

```{r}
#| label: solve-system
# System: 2x + y = 5, x + 3y = 7
A <- matrix(c(2, 1, 1, 3), nrow = 2, byrow = TRUE)
b <- c(5, 7)

x <- solve(A, b)
x

# Verify
A %*% x
```

### Application: mixed Nash equilibrium

In a 2x2 game, finding the mixed Nash equilibrium reduces to making each player indifferent between their pure strategies. If Player 2 mixes with probability $q$ on the first action, Player 1's indifference condition is:

$$
a_{11} q + a_{12}(1-q) = a_{21} q + a_{22}(1-q)
$$

This is a single linear equation in $q$. For larger games, mixed equilibria require solving systems of such equations simultaneously.

```{r}
#| label: mixed-ne-linalg
# Prisoner's Dilemma payoffs for Player 1
# Row player payoffs: A[i,j]
A <- matrix(c(3, 0, 5, 1), nrow = 2, byrow = TRUE)

# Indifference: A[1,1]*q + A[1,2]*(1-q) = A[2,1]*q + A[2,2]*(1-q)
# Rearranging: q*(A[1,1] - A[1,2] - A[2,1] + A[2,2]) = A[2,2] - A[1,2]
denom <- A[1, 1] - A[1, 2] - A[2, 1] + A[2, 2]
q_star <- (A[2, 2] - A[1, 2]) / denom
q_star
```

## Eigenvalues and eigenvectors

An eigenvector $\mathbf{v}$ of a square matrix $A$ satisfies $A\mathbf{v} = \lambda \mathbf{v}$ for some scalar $\lambda$ (the eigenvalue). Eigenvalues appear in several places in this book:

- **Replicator dynamics** (@sec-replicator-dynamics): the stability of fixed points is determined by the eigenvalues of the Jacobian matrix.
- **Markov chains**: the stationary distribution is the eigenvector corresponding to eigenvalue 1.

```{r}
#| label: eigen
M <- matrix(c(4, 1, 2, 3), nrow = 2, byrow = TRUE)

eig <- eigen(M)
eig$values
eig$vectors

# Verify: M v = lambda v
lambda_1 <- eig$values[1]
v_1 <- eig$vectors[, 1]
M %*% v_1
lambda_1 * v_1
```

### Stability and the Jacobian

For a dynamical system $\dot{\mathbf{x}} = f(\mathbf{x})$, a fixed point $\mathbf{x}^*$ is **stable** if all eigenvalues of the Jacobian matrix $J = \partial f / \partial \mathbf{x}$ evaluated at $\mathbf{x}^*$ have negative real parts.

```{r}
#| label: jacobian-stability
# Example: Jacobian at a fixed point of a 2D system
J <- matrix(c(-2, 1, 0, -3), nrow = 2, byrow = TRUE)

eig_J <- eigen(J)
eig_J$values

# Both eigenvalues are negative => stable fixed point
all(Re(eig_J$values) < 0)
```

## Positive definite matrices

A symmetric matrix $A$ is **positive definite** if $\mathbf{x}^T A \mathbf{x} > 0$ for all nonzero $\mathbf{x}$. Equivalently, all eigenvalues are positive. Positive definiteness arises in quadratic payoff functions and in verifying second-order conditions for optimisation problems.

```{r}
#| label: pos-def
A <- matrix(c(4, 1, 1, 3), nrow = 2, byrow = TRUE)

# Check via eigenvalues
eigen(A)$values
all(eigen(A)$values > 0)

# Check via Cholesky decomposition (succeeds only for positive definite matrices)
chol(A)
```

## Convex sets and the simplex

### The probability simplex

A **mixed strategy** is a probability distribution over pure strategies. The set of all such distributions forms the **(probability) simplex**:

$$
\Delta_n = \left\{ \mathbf{p} \in \mathbb{R}^n \;\middle|\; p_i \geq 0,\; \sum_{i=1}^{n} p_i = 1 \right\}
$$

For $n = 2$, the simplex is a line segment; for $n = 3$, it is a triangle.

```{r}
#| label: simplex-check
# Check whether a vector is a valid mixed strategy
is_mixed_strategy <- function(p, tol = 1e-10) {
  all(p >= -tol) && abs(sum(p) - 1) < tol
}

is_mixed_strategy(c(0.3, 0.5, 0.2))
is_mixed_strategy(c(0.5, 0.6, -0.1))
```

### Convex sets and convex combinations

A set $C$ is **convex** if for any two points $\mathbf{x}, \mathbf{y} \in C$ and any $\lambda \in [0, 1]$, the point $\lambda \mathbf{x} + (1 - \lambda) \mathbf{y}$ is also in $C$. Important convex sets in game theory include:

- The simplex $\Delta_n$ (the set of mixed strategies).
- The **core** of a cooperative game (see @sec-cooperative-gt).
- The **feasible payoff region** of a repeated game (see @sec-repeated-games).

```{r}
#| label: convex-combination
# Convex combination of two strategy vectors
p1 <- c(0.7, 0.2, 0.1)
p2 <- c(0.1, 0.5, 0.4)
lambda <- 0.6

p_mix <- lambda * p1 + (1 - lambda) * p2
p_mix
is_mixed_strategy(p_mix)
```

### The support of a mixed strategy

The **support** of a mixed strategy $\mathbf{p}$ is the set of pure strategies that receive positive probability: $\text{supp}(\mathbf{p}) = \{i : p_i > 0\}$. In a Nash equilibrium, every pure strategy in the support must yield the same expected payoff (the indifference principle from @sec-mixed-strategies).

```{r}
#| label: support
support <- function(p, tol = 1e-10) {
  which(p > tol)
}

support(c(0.5, 0, 0.5, 0))
support(c(0.25, 0.25, 0.25, 0.25))
```

## Matrix decompositions

### LU decomposition

R does not expose LU decomposition directly in base, but `solve()` uses it internally. For most game-theoretic applications, `solve()` suffices.

### Singular value decomposition (SVD)

The SVD decomposes any $m \times n$ matrix as $A = U \Sigma V^T$. It is useful for low-rank approximations and for diagnosing near-singular payoff matrices.

```{r}
#| label: svd
A <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, byrow = TRUE)

s <- svd(A)
s$d  # singular values

# Reconstruct: U %*% diag(d) %*% t(V) = A
s$u %*% diag(s$d) %*% t(s$v)
```

## Summary of R linear algebra functions

| Task | R function | Example |
|:-----|:-----------|:--------|
| Matrix multiply | `%*%` | `A %*% B` |
| Transpose | `t()` | `t(A)` |
| Determinant | `det()` | `det(A)` |
| Inverse | `solve()` | `solve(A)` |
| Solve $Ax = b$ | `solve()` | `solve(A, b)` |
| Eigenvalues | `eigen()` | `eigen(A)$values` |
| SVD | `svd()` | `svd(A)$d` |
| Cholesky | `chol()` | `chol(A)` |
| Cross product | `crossprod()` | `crossprod(A)` gives $A^T A$ |
| Outer product | `outer()` or `%o%` | `x %o% y` |

: R functions for linear algebra {#tbl-linalg-functions}
